Chain level Steenrod operations

### 2021-05-07 groups.tex
Split sections into files

\documentclass[A4]{amsart}

\usepackage{amsmath, amssymb, amsfonts}
\usepackage{tikz-cd}
\usepackage{graphics}
\usepackage[bookmarks=true, linktocpage=true,
bookmarksnumbered=true, breaklinks=true,
pdfstartview=FitH, hyperfigures=false,
plainpages=false, naturalnames=true,
colorlinks=true, pagebackref=true,
pdfpagelabels]{hyperref}
\hypersetup{
	colorlinks,
	citecolor=blue,
	filecolor=blue,
	linkcolor=blue,
	urlcolor=blue
}
\usepackage[alphabetic]{amsrefs}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{remark}[theorem]{Remark}

\newcommand{\Z}{\mathbb{Z}}
\newcommand{\Hom}{\mathrm{Hom}}
\newcommand{\End}{\mathrm{End}}
\DeclareMathOperator*{\colim}{colim}

\include{figures}

\setlength{\textwidth}{\paperwidth}
\addtolength{\textwidth}{-1.6in}
\setlength{\textheight}{\paperheight}
\addtolength{\textheight}{-1.8in}
\calclayout

\begin{document}

\title{Chain level Steenrod operations}

\author[R. Kaufmann]{Ralph M.\ Kaufmann}
\address{Department of Mathematics, Department of Physics and Astronomy, Purdue University}
\email{rkaufman@purdue.edu}

\author[A. Medina-Mardones]{Anibal M. Medina-Mardones}
\address{Laboratory for Topology and Neuroscience, École Polytechnique Fédérale de Lausanne, Switzerland}
\address{Department of Mathematics, University of Notre Dame du Lac, Notre Dame, IN, USA}
\email{amedinam@nd.edu}
\thanks{A.M-M. acknowledges financial support from Innosuisse grant \mbox{32875.1 IP-ICT - 1}, and the hospitality of the Max Plank Institute for Mathematics.}

\subjclass[2010]{55S05, 55U05, 55S10, 55S12; 55-04, 55U15, 55S15}
\keywords{Steenrod operations, Dyer-Lashof operations, simplicial cochains, cubical cochains}

\begin{abstract}
	Steenrod defined in 1947 the Steenrod squares on the mod 2 cohomology of spaces using explicit cochain formulae for the cup-$i$ products; a family of coherent homotopies derived from the broken symmetry of Alexander-Whitney's chain approximation to the diagonal. Later, following a viewpoint developed by Adem, Steenrod defined his homonymous operations for all primes using the homology of symmetric groups. This approach enhanced the conceptual understanding of the operations and allowed for many advances but lacked the concreteness of their definition at~$2$. In recent years, thanks to the development of new applications of cohomology ---most notably in Applied Topology and Quantum Field Theory--- having a definition of Steenrod operations that can be effectively computed in specific examples has become a key issue. This article provides such definition, at all primes, for well known $E_\infty$-algebras including the cochains of spaces expressed via simplicial or cubical cell structures.
\end{abstract}

\maketitle

\tableofcontents

\section{Introduction}

The role Steenrod operations play in Stable Homotopy Theory is hard to overstate. The reason is that, given the representability of the cohomology functor, these operations together with the Bockstein homomorphism can be used to give a complete description of the algebraic structure naturally present on the mop-$p$ \mbox{cohomology} algebra of spaces. For the even prime, Steenrod squares were introduced in \cite{steenrod47products} via an explicit choice of coherent homotopical corrections to the broken symmetry of Alexander-Whitney's chain approximation to the diagonal. Later, for odd primes,	their definition was given non-effectively using arguments based on the mod-$p$ homology of symmetric groups \cites{steenrod53symmetric, steenrod53cyclic, steenrod62operations}. This viewpoint enhanced the conceptual understanding of the operations and allowed for many advances \cites{adem1952iteration, milnor1958dual, adams1995stable}, but lacked the concreteness of their definition at~$2$.

In recent years, thanks to the development of new applications of cohomology ---most notably in Applied Topology and Quantum Field Theory--- the need to have a definition of Steenrod operations that can be effectively computed in specific examples has gain considerable importance. In Applied Topology, the use of Persistence Homology \cites{carlsson2009data, edelsbrunner2008persistent} has open many new interdisciplinary research directions \cites{de2007coverage,chan2013topology,lee2017quantifying}, and the availability of formulae for the cup-$i$ products allowed for the development of a theory of Persistence Steenrod modules accessing finer features of the data \cite{medina2018persistence}. In Quantum Field Theory, the BRST formalism \cites{BRS,Tyutin,KugoOjima} introduced cohomology to deal with gauge symmetries, and was further developed in the BV-formalism \cite{BV}. Master equations, introduced there, are also needed for the construction of viable actions in String Field Theory \cite{Zwiebach}, where fundamental chains are involved. The chain level viewpoint for the study of homology operations was brought to the forefront by String Topology \cite{Sullivanoverview}, and as served as a major driver of innovation \cites{TZ, hoch2}. The cubical viewpoint appears in physics naturally in the theory of renomalization \cite{KreimerCubical}; while Steenrod cup-$i$ products are utilized in the construction of effective actions of gapped systems \cite{kapustin2017fermionic}. This connection with physics motivated the development of effective versions of spin bordism \cites{brumfiel2016pontrjagin, brumfiel2018pontrjagin}, which prominently feature Steenrod cup-$i$ products and Cartan coboundaries.

As in \cite{may70generalapproach}, we take a more general approach to Steenrod operations that also includes Araki-Kudo-Dyer-Lashof operations on the mod-$p$ homology of infinite loop spaces \cites{araki56squaring, dyer62lashof}. We use the language of operads \cite{may72geometry}  to describe at the chain level the integral structure required to define (co)homology operations at every prime. We then describe effective constructions of this structure on three prominent models of the $E_\infty$-operad, identifying elements in them that represent Steenrod operations in the mod-$p$ homology of their algebras; these are the Barratt-Eccles \cite{berger04combinatorial}, surjection \cite{mcclure03cochain}, and $U(\mathcal M)$ \cite{medina2020prop1} operads. Since the cochains of simplicial sets are equipped with effective and compatible algebra structures over each of these operads, we are able to explicitly describe canonical products generalizing the \mbox{cup-$i$} products of Steenrod to every prime. An alternative approach based on the Eilenberg-Zilber contraction can be found in \cite{gonzalez2005hpt}. Up to now, although its existence is warrantied by acyclic model argument \cite{eilenberg1953acyclic}, no $E_\infty$-structure was known on the cochains of cubical sets. Using an $E_\infty$-bialgebra structure on the chains of the interval we effectively construct an $U(\mathcal M)$-algebra structure on them, and as a consequence provide a description of Steenrod products for cochains of cubical sets as well. The $p=2$ part of our constructions specialize in the simplicial context to the definition of cup-$i$ products given by Steenrod \cite{steenrod47products} and recovered by several authors \cites{mcclure03cochain, berger04combinatorial, medina2018axiomatic}, and in the cubical setting to the cup-$i$ products of  Kadeishvili \cite{kadeishvili1998dg} and Kr\v{c}\'{a}l-Pilarczy \cite{pilarczyk2016cubical}.

We also contribute to a growing body of open source software \cites{manero2020effective, pilarczyk2015contraction, sagemath} resulting from an effort to rewrite key concepts from Homotopy Theory in effective forms. Uses in Applied Topology are motivating some of this work \cite{tauzin2020giottotda}, but the impetus also comes from Voevodsky's program for a comprehensive, computational foundation of mathematics based on the homotopical interpretation of type theory \cite{bauer2017hott}. A \texttt{Python} implementation of all the constructions in this article can be found in A.M-M.'s website.\footnote{Currently hosted at \url{https://www.medina-mardones.com/}}

An outline of the article is described next.
We first introduce, in Section~2, the conventions we will follow regarding chain complexes, simplicial sets and cubical sets.
Then, in Section~3, we review the key notions from group homology we will use for cyclic and symmetric groups. Section~4 is devoted to the language of operads and related structures that we use, in Section~5, to define an integral structure at the chain level inducing Steenrod operations for every prime. Section~6, the bulk of this work, presents effective constructions of Steenrod-Adem structures on the Barratt-Eccles, surjection, and $U(\mathcal M)$ operads.
It also describes a natural $U(\mathcal M)$-algebra structure on the cochains of simplicial and cubical sets inducing natural Steenrod-Adem structures on them.
We end, in Section~7, with an overview of some connections of this work to important geometric and combinatorial structures, and provide an outline of future research directions.

\section*{Acknowledgement}

The authors thank Dennis Sullivan, Kathryn Hess, Clemens Berger, John Morgan, Greg Brumfiel, Dev Sinha, Jens Kjaer, Calista Bernard, and Andy Putnam, for insightful discussions and valuable comments, and to the anonymous referee for many helpful suggestions.

\section{Preliminaries}

\subsection{Chain complexes}
\label{ssec: chain cpx}

Let $R$ be a ring. We denote by $(\mathbf{Ch}_R, \otimes, R)$ the symmetric monoidal category of homologically graded chain complexes of $R$-modules. The set of $R$-linear maps between chain complexes as well as the tensor product of chain complexes are regarded as chain complexes in the usual way:
\begin{equation*}
\Hom(A, A^\prime)_n = \big\{f \ |\ a \in A_m \Rightarrow f(a) \in A^\prime_{m+n} \big\}, \qquad
\partial f = \partial \circ f - (-1)^{|f|}f \circ \partial,
\end{equation*}
\begin{equation*}
(A \otimes A^\prime)_n = \bigoplus_{p + q = n} A_p \otimes A^\prime_q, \qquad \partial (a \otimes a^\prime) = \partial a \otimes a^\prime + (-1)^{|a|} a \otimes \partial a^\prime.
\end{equation*}
We embed the category of $R$-modules as the full subcategory of $\mathbf{Ch}_R$ with objects concentrated in degree $0$. The endofunctor $\Hom(-, R)$ is referred to as the \textit{linear duality functor}. We notice that if a chain complex is concentrated in non-negative degrees then its linear dual is on non-positive ones.

The rings we will mostly be interested in are the group rings $\Z[\mathrm G]$ and $\mathbb{F}_p[\mathrm G]$ of finite groups, where $p$ is prime and $\mathbb{F}_p$ is the field with $p$ elements.

\subsection{Simplicial sets}

The \textit{simplex category} $\triangle$ is defined to have an object $[n] = \{0, \dots, n\}$ for every $n \in \mathbb{N}$ and a morphism $[m] \to [n]$ for each order-preserving function from $[m]$ to $[n]$. The morphisms $\delta_i : [n-1] \to [n]$ and $\sigma_i : [n+1] \to [n]$ defined for $0 \leq i \leq n$ by
\begin{equation*}
\delta_i(k) = 
\begin{cases} k & k < i, \\ k+1 & i \leq k, \end{cases}
\quad \text{ and } \quad
\sigma_i(k) = 
\begin{cases} k & k \leq i, \\ k-1 & i < k, \end{cases}
\end{equation*} 
generate all morphisms in the simplex category.

A \textit{simplicial set} $X$ is a contravariant functor from the simplex category to the category of sets and a simplicial map is a natural transformation between two simplicial sets. As is customary we use the notation
\begin{equation*}
X\big( [n] \big) = X_n, \qquad X(\delta_i) = d_i, \qquad X(\sigma_i) = s_i,
\end{equation*}
and refer to elements in the image of any $s_i$ as \textit{degenerate}.

For each $n \in \mathbb{N}$, the simplicial set $\triangle^n$ is defined by
\begin{equation*}
\triangle^n_k  = \Hom_{\triangle} \big( [k], [n] \big), \qquad 
d_i(x) = x \circ \delta_i, \qquad 
s_i(x) = x \circ \sigma_i,
\end{equation*}
and any simplicial set can be expressed as a colimit of these
\begin{equation*}
X = \colim_{\triangle^n \to X} \triangle^n.
\end{equation*}
We represent the non-degenerate elements of $\triangle^n_k$ as an increasing sequence $[v_0, \dots, v_k]$ of non-negative integers each less than or equal to $n$.

The functor $N_\bullet$ of \textit{normalized chains} (with $R$-coefficients) is defined as follows: 
\begin{equation*}
N_\bullet(X; R)_n = \frac{R \{ X_n \}}{R \{ s(X_{n-1}) \}}
\end{equation*}
where $s(X_{n-1}) = \bigcup_{i=0}^{n-1} s_i(X_{n-1})$, and $\partial_n : N_*(X)_n \to N_*(X)_{n-1}$ is given by
\begin{equation*}
\partial_n = \sum_{i=0}^{n} (-1)^id_{i}.
\end{equation*}
The functor of \textit{normalized cochains} $N^\bullet$ is defined by composing $N_\bullet$ with the linear duality functor $\Hom(-, R)$.

It is convenient to emphasize that
\begin{equation*}
N_\bullet(X; R) = \colim_{\triangle^n \to X} N_\bullet(\triangle^n; R).
\end{equation*}

\subsection{Cubical sets}

The \textit{cube category} $\square$ is the free strict monoidal category with a \textit{bipointed object}
\begin{equation*}
\begin{tikzcd}
1 \arrow[r, bend left, "\delta^0"] \arrow[r, bend right, "\delta^1"'] & 2 \arrow[r, "\sigma"] & 1
\end{tikzcd}
\end{equation*}
such that $\sigma \circ \delta^0 = \sigma \circ \delta^1 = \mathrm{id}$. Explicitly, it contains an object $2^n$ for each non-negative integer $n$ and its morphisms are generated by the \textit{coface} and \textit{codegeneracy maps} defined by
\begin{align*}
\delta_i^\varepsilon & = \mathrm{id}_{2^{i-1}} \times \delta^\varepsilon \times \mathrm{id}_{2^{n-1-i}} \colon 2^{n-1} \to 2^n, \\
\sigma_i & = \mathrm{id}_{2^{i-1}} \times \, \sigma \times \mathrm{id}_{2^{n-i}} \colon 2^{n} \to 2^{n-1}.
\end{align*}

A \textit{cubical set} $X$ is a contravariant functor from the cube category to the category of sets and a cubical map is a natural transformation between two cubical sets. As is customary, we use the notation
\begin{equation*}
X\big( 2^n \big) = X_n \qquad X(\delta^\varepsilon_i) = d^\varepsilon_i \qquad X(\sigma_i) = s_i,
\end{equation*}
and refer to elements in the image of any $s_i$ as \textit{degenerate}.

For each $n \in \mathbb{N}$, the cubical set $\square^n$ is defined by
\begin{equation*}
\square^n_k  = \Hom_{\square} \big( 2^k, 2^n \big), \qquad 
d^\varepsilon_i(x) = x \circ \delta^\varepsilon_i, \qquad 
s_i(x) = x \circ \sigma_i,
\end{equation*}
notice that iteratively
\begin{equation*}
\square^n = \overbrace{\square^1 \times \cdots \times \square^1}^{n \text{ times }}.
\end{equation*}
We represent the non-degenerate elements of $\square^n$ as sequences $x_1 \cdots\, x_n$ with each $x_i \in \{[0], [1], [0,1]\}$. Any cubical set can be expressed as a colimit of these
\begin{equation*}
X = \colim_{\square^n \to X} \square^n.
\end{equation*}

The functor $N_\bullet$ of \textit{normalized chains} (with $R$-coefficients) is defined as follows: The chain complex $N_\bullet(\square^1)$ is simply the cellular chain complex of the interval,
isomorphic to
\begin{equation*}
\begin{tikzcd} [column sep = small, row sep = 0.1pt]
R\{[0], [1]\}  & \arrow[l] R\{[0,1]\} \\
{[1] - [0]} & \arrow[l, |->] \left[0,1\right].
\end{tikzcd}
\end{equation*}
Set
\begin{equation} \label{eq: chains on I^n}
N_\bullet(\square^n; R) = N_\bullet(\square^1; R)^{\otimes n}
\end{equation}
and define
\begin{equation*}
N_\bullet(X; R) = \colim_{\square^n \to X} N_\bullet(\square^n; R).
\end{equation*}

The functor of \textit{normalized cochains} $N^\bullet$ is defined by composing $N_\bullet$ with the linear duality functor $\Hom(-, R)$.

\section{Group homology}
\label{sec: goup homology}

@@

\section{$\Gamma$-modules, operads and PROPs}

In this section we set up a framework in which the structure responsible for Steenrod operations becomes most transparent. Given our applications, we consider $\mathbf{Ch}_R$ as the base category, remarking that all definitions in this section apply to general closed symmetric monoidal categories.

\subsection{$\Gamma$-modules}
Recall that a group $\mathrm G$ can be thought of as a category with a single object and only invertible morphisms, and that a chain complex of left (resp. right) $R[\mathrm G]$-modules is the same as a covariant (resp. contravariant) functor from $\mathrm G$ to $\mathbf{Ch}_R$. Right actions are more commonly used in the literature of operads so we emphasize that perspective. Taking inverses allows for the switch between these two conventions.

A \textit{groupoid} is a small category where all morphisms are invertible.
\begin{definition}
	A \textit{$\Gamma$-module} is a contravariant functor to $\mathbf{Ch}_R$ from a groupoid $\Gamma$ with objects the natural numbers and morphisms satisfying $\Gamma(r,s) = \emptyset$ for $r \neq s$. We denote the category of $\Gamma$-modules and natural transformations by $\mathbf{Ch}_R^\Gamma$.
\end{definition}

We are mostly interested in two examples of $\Gamma$-modules, those associated to the groupoids $\mathrm{S}$ and $\mathrm{C}$ defined by
\begin{equation*}
\mathrm{S}(r, r) = \mathrm{S}_r, \qquad
\mathrm{C}(r,r) = \mathrm{C}_r,
\end{equation*}
for every $r \in \mathbb{N}$. The inclusion $\mathrm{C}_r \to \mathrm{S}_r$ induces a forgetful functor
\begin{equation*}
\begin{tikzcd} [column sep = small]
\mathbf{Ch}_R^\mathrm{S}  \arrow[r] & \mathbf{Ch}_R^\mathrm{C}.
\end{tikzcd}
\end{equation*}

Given an object $A$ in $\mathbf{Ch}_R$ there are two important $\Gamma$-modules associated to it; an $\mathrm{S}$-module known as \textit{endomorphism $\mathrm{S}$-module} $\End_A$, and an $\mathrm{S}^{op}$-module known as \textit{coendomorphism $\mathrm{S}$-module} $\End^A$. These are defined by
\begin{align*}	
\End_A(r) &= \Hom(A^{\otimes r},A), \\
\End^A(r) &= \Hom(A,A^{\otimes r}),
\end{align*}
with respective right and left actions defined by permutation of tensor factors.

Another groupoid of importance to us is $\mathrm{S} \times \mathrm{S}^{op}$. Contravariant functors from it to $\mathbf{Ch}_R$ are called \textit{$\mathrm{S}$-bimodules} and their functor category is denoted $\mathbf{Ch}_R^{\mathrm{S} \times \mathrm{S}^{op}}$. Notice that the inclusions
define forgetful functors
\begin{equation*}
\begin{tikzcd}[column sep=small, row sep=tiny]
& \mathbf{Ch}_R^{\mathrm{S} \times \mathrm{S}^{op}} \arrow[dl, "U_1"'] \arrow[dr, "U_2"] & \\
\mathbf{Ch}_R^{\mathrm{S}} &  & \mathbf{Ch}_R^{\mathrm{S}^{op}},
\end{tikzcd}
\end{equation*}
and that for any object $A$ in $\mathbf{Ch}_R$, the canonical \textit{endomorphism bimodule}
\begin{equation*}
\End_A^A(r, s) = \Hom(A^{\otimes r}, A^{\otimes s})
\end{equation*}
forgets to $\End_A$ and $\End^A$.

Using the groupoid automorphism sending every morphisms to its inverse we can identify $\Gamma$- and $\Gamma^{op}$- modules. In particular, the statement claiming that the linear duality functor induces a morphism of $\mathrm{S}$-modules
\begin{equation*}
\End^A \to \End_{\Hom(A, R)}
\end{equation*}
for every object $A$ in $\mathbf{Ch}_R$ is well defined and easily verified using this identification.

A \textit{resolution} in $\mathbf{Ch}_R^\Gamma$ is a morphism $\phi$ of $\Gamma$-modules such that $\phi(r)$ is a resolution in the category of chain complexes of $R[\Gamma_r]$-modules for each $r \in \mathbb{N}$. A $\Gamma$-module $\mathcal R$ is said to be $E_\infty$ if $\mathcal R(0) = R$ and there exists a resolution $\mathcal R \to \underline{R}$ where $\underline{R}$ is the object in $\mathbf{Ch}_R^\Gamma$ defined by $\underline{R}(r) = R$ and $\underline{R}(\gamma) = \mathrm{id}_R$ for every $r \in \mathbb{N}$ and $\gamma \in \Gamma_r$.

\begin{definition} \label{def: minimal cyclic resolution}
	The \textit{minimal} $E_\infty$-$\mathrm{C}$-\textit{module} $\mathcal W$ is the functor in $\mathbf{Ch}_R^\mathrm{C}$ assigning to $r$ the chain complex
	\begin{equation*}
	\begin{tikzcd} [column sep = .5cm]
	R[\mathrm{C}_r] & \arrow[l, "\,T"'] R[\mathrm{C}_r] & \arrow[l, "\,N"'] R[\mathrm{C}_r] & \arrow[l, "\,T"'] \cdots
	\end{tikzcd}
	\end{equation*}
	concentrated in non-negative degrees.
\end{definition}

\subsection{Operads and PROPs}

Operads an PROPs are respectively $\mathrm{S}$-modules and \mbox{$\mathrm{S}$-bimodules} enriched with further compositional structure. These structures are best understood by abstracting the compositional structure naturally present in the endomorphism $\mathrm{S}$-module $\End_A$, naturally an operad, and the endomorphism $\mathrm{S}$-bimodule $\End_A^A$, naturally a PROP.

Succinctly, an operad $\mathcal O$ is an $\mathrm{S}$-module together with a collection of $R$-linear maps
\begin{equation*}
\mathcal O(r) \otimes \mathcal O(s) \to \mathcal O(r+s-1)
\end{equation*}
satisfying suitable associativity, equivariance and unitality conditions. A PROP $\mathcal P$ is an $\mathrm{S}$-bimodule together with two types of compositions; horizontal
\begin{equation*}
\mathcal P(r_1, s_1) \otimes \mathcal P(r_2, s_2) \to \mathcal P(r_1 + r_2, s_1 + s_2)
\end{equation*}
and vertical
\begin{equation*}
\mathcal P(r,s) \otimes \mathcal P(s, t) \to \mathcal P(r, t)
\end{equation*}
satisfying their own versions of associativity, equivariance and unitality. For a complete presentation of these concepts we refer to Definition 11 and 54 of \cite{markl2008operads}.

We have already mentioned that for any chain complex $A$, the $\mathrm{S}$-module $\End_A$ is naturally an operad and the $\mathrm{S}$-bimodule $\End_A^A$ is naturally a PROP. We add that considering $\End^A$ as an $\mathrm{S}$-module, its compositional structure also defines an operad structure on it.

We remark that the compositional structure of a PROP $\mathcal P$ restricts to operad structures on $U_1(\mathcal P)$ and $U_2(\mathcal P)$.

The type of operads that we are most interested in are $E_\infty$-operads which, as we will discuss in the next section, are used to describe commutativity up to coherent homotopies.

\begin{definition} [\cite{may72geometry}, \cite{boardman1973homotopy}]
	An operad is said to be $E_\infty$ if its underlying \mbox{$\mathrm{S}$-module} is $E_\infty$, and a PROP $\mathcal P$ is said to be $E_\infty$ if either $U_1(\mathcal P)$ or $U(\mathcal P)$ is an \mbox{$E_\infty$-operad}.
\end{definition}

\subsection{Algebras, coalgebra and bialgebras}

A morphisms of operads or of PROPs is simply a morphisms of their underlying $\mathrm{S}$-modules or $\mathrm{S}$-bimodules preserving the respective compositional structures.

Given a chain complex $A$, an operad $\mathcal O$ and a PROP $\mathcal P$. An $\mathcal O$-\textit{algebra} (resp. $\mathcal O$-\textit{coalgebra}) structure on $A$ is an operad morphism $\mathcal O \to \End_A$ (resp. $\mathcal O \to \End^A$), and a $\mathcal P$-\textit{bialgebra} structure on $A$ is a PROP morphism $\mathcal P \to \End_A^A$.

We remark that the linear duality functor naturally transforms an $\mathcal O$-coalgebra structure on a chain complex into an $\mathcal O$-algebra structure on its dual.

Algebras over $E_\infty$-operads are the central objects of study in this work. To develop intuition for them, let us consider a chain complex $A$ with an algebra structure over $\underline{R}$, thought of as an operad with all compositions corresponding to the identity map $R \to R$. The $\underline{R}$-algebra structure on $A$ is generated by a linear map $\mu \colon A \otimes A \to A$ which is (strictly) commutative and associative, and a linear map $\eta \colon R \to A$ that determines a (two-sided) unit for $\mu$. Since $E_\infty$-operads are resolutions of $\underline{R}$, their algebras can be thought of as usual unital algebras where the commutativity and associativity relations hold up to coherent homotopies. The two main examples to keep in mind are the cochains of spaces and the chains of infinite loops spaces.

\section{Steenrod-Adem structures}

We now introduce an operadic formulation of a structure giving rise to Steenrod operations at every prime. It is based on \cites{steenrod47products, steenrod53symmetric, steenrod53cyclic, may70generalapproach}, and takes a viewpoint that emphasizes the integral structure needed to define them and prove their relations at every prime. For a more geometric treatment we refer the reader to \cites{may72geometry, may76homology, lawson2020n}.

\begin{definition} \label{def: Steenrod-Adem structure}
	A \textit{Steenrod-Adem structure} on an operad $\mathcal O$ is a 
	morphism of $\mathrm{C}$-modules $\psi \colon \mathcal W \to \mathcal O$ for which there exists a factorization through an $E_\infty$-operad
	\begin{equation*}
	\begin{tikzcd}[column sep = normal, row sep = small]
	& \mathcal R \arrow[dr, "\phi", out=0] & \\
	\mathcal W \arrow[ur, "\iota", in=180] \arrow[rr, "\psi"] & & \mathcal O
	\end{tikzcd}
	\end{equation*}
	such that $\iota$ is a quasi-isomorphism and $\phi$ a morphism of operads.
\end{definition}

\begin{remark} \label{rmk: Deligne conjecture}
	In \cite{GerstenhaberVoronov} an operad morphism $\mathcal{A}ssoc \to \mathcal O$ is referred to as a \textit{multiplication} on $\mathcal O$. In this language, a choice of factorization for a Steenrod-Adem structure defines in particular a \textit{commutative multiplication}.
\end{remark}

\begin{definition}
	Let $A$ be a chain complex. A Steenrod-Adem structure on $\mathrm{End}_A$ is referred to as one on $A$. Given one such structure $\psi \colon \mathcal W \to \End_A$, the \textit{Steenrod} $(r, i)$-\textit{product} of $A$ is defined for every $r, i \geq 0$ as the image in $ \mathrm{End}(A^{\otimes r}, A)$ of $\psi(e_i)$.
\end{definition}

\begin{remark}
	Restricting a Steenrod-Adem structure on a chain complex $A$ to a given arity $r$ corresponds to specifying a map $\theta \colon \mathcal W(r) \otimes A^{\otimes r} \to A$. This pair $(A, \theta)$ is an object in May's category $\mathfrak{C}(\mathrm C_r, \infty, \Z)$ introduced in \cite{may70generalapproach} Definition 2.1. Furthermore, the restriction to arity $r^2$ enhances this pair to an \textit{Adem object} as described in Definition 4.1 loc. cit.
\end{remark}

A Steenrod-Adem structure on a chain complex $A$ allows us to define operations on its $\mathbb{F}_p$-homology for every prime $p$. These operations are structurally different if $p$ is even or odd. Using the previous Remark, the definitions below recover Steenrod's original definition for the cohomology of spaces and that of Araki-Kudo-Dyer-Lashof operations for the homology of infinite loop spaces.

For the rest of this section $A$ is a chain complex equipped with a Steenrod-Adem structure.

\begin{definition}
	For any prime $p$, the $\mathbb{F}_p$-linear map
	\begin{equation*}
	D^p_i : (A \otimes \mathbb{F}_p) \to (A \otimes \mathbb{F}_p)
	\end{equation*}
	is defined by sending $a$ to the Steenrod \mbox{$(p, i)$-product} of $(a \otimes \cdots \otimes a) \in (A \otimes \mathbb{F}_p)^{\otimes p}$ if $i \geq 0$ and to $0$ otherwise.
\end{definition}

We notice that if $a$ is of degree $q$ then $D^p_i(a)$ is of degree $q + (p-1)q + i$.

\begin{definition}
	For any integer $s$, the \textit{Steenrod operation}
	\begin{equation*}
	P_s : H_\bullet(A; \mathbb{F}_2) \to H_{\bullet + s}(A; \mathbb{F}_2)
	\end{equation*}
	is defined by sending the class represented by a cycle $a \in (A \otimes \mathbb{F}_2)$ of degree $q$ to the class represented by $D^2_{s-q}(a)$.
\end{definition}

Notice that the Steenrod squares are determined by the Steenrod $(2,i)$-products with $\mathbb{F}_2$-coefficients. These binary operations are known as cup-$i$ products. In a similar way, the operations $P$ and $\beta P$ defined below for odd primes are determined by the Steenrod $\big(p, k(p-1)-\varepsilon\big)$-products for $\varepsilon \in \{0,1\}$. We can explain the appearance of these specific Steenrod products as follows. The increase on the degree of a $q$-cycle after applying $D^p_{k(p-1)-\varepsilon}$ to it is $(p-1)(q+k) - \varepsilon$, which can be rewritten as $2t(p-1) - \varepsilon$ if $q$ is even an $(2t+1)(p-1) - \varepsilon$ if $q$ is odd. According to Lemma \ref{lem: Thom's theorem}, these are the only homologically non-trivial cases.

\begin{definition} \label{def: Steenrod operations at odd prime}
	For any integer $s$, the \textit{Steenrod operations}
	\begin{equation*}
	P_s : H_\bullet(A; \mathbb{F}_p) \to H_{\bullet + 2s(p-1)}(A; \mathbb{F}_p)
	\end{equation*}
	and
	\begin{equation*}
	\beta P_s : H_\bullet(A; \mathbb{F}_p) \to H_{\bullet + 2s(p-1) - 1}(A; \mathbb{F}_p)
	\end{equation*}
	are defined by sending the class represented by a cycle $a \in (A \otimes \mathbb{F}_p)$ of degree $q$ to the classes represented respectively for $\varepsilon \in\{0,1\}$ by
	\begin{equation*}
	(-1)^s \nu(q) D^p_{(2s-q)(p-1)-\varepsilon}(a)
	\end{equation*}
	where $\nu(q) = (-1)^{q(q-1)m/2}(m!)^q$ and $m = (p-1)/2$.
\end{definition}

\begin{remark}
	The use of the coefficient function $\nu(q)$ is motivated by the identity $D_{q(p-1)}^p(a) = \nu(q)a$ in the case of spaces (see Eq. 6.1. in \cite{steenrod53cyclic}). The notation $\beta P_s$ is motivated by the relationship of this operator and the Bockstein of the reduction $\mathbb Z \to \mathbb Z/p\mathbb Z$.
\end{remark}

Steenrod operations satisfy the Cartan and Adem relations. Below we state the subset of these relations that is most commonly used and refer to Sections 2 and 4 of \cite{may70generalapproach} for a complete list and proofs.
\begin{lemma}
	Let $A$ be equipped with a Steenrod-Adem structure. If $[\alpha], [\beta] \in H(A;\mathbb{F}_p)$ then
	\begin{equation*}
	P_s\big([\alpha] [\beta]\big) = \sum_{i+j=s} P_i\big( [\alpha] \big) P_j\big( [\beta] \big),
	\end{equation*}
	and
	\begin{itemize}
		\item If $p = 2$ and $a > 2b$
		\begin{equation*}
		P_{a}P_{b} = \sum_i {2i-a \choose a-b-i-1} P_{a+b-i}P_i,
		\end{equation*}
		\item If $p > 2$ and $a > pb$
		\begin{equation*}
		P_{a}P_{b} = \sum_i (-1)^{a+i}{pi-a \choose a-(p-1)b-i-1} P_{a+b-i}P_i.
		\end{equation*}
	\end{itemize}		
\end{lemma}

\begin{proof}
	These are proven as part of Corollary 2.7 and Theorem 4.7 in \cite{may70generalapproach}.
\end{proof}

For the even prime, proofs at the chain level of these formulae have been given in \cite{medina2020cartan,brumfiel2020cochain}. For a different operadic approach to Steenrod operations at the even prime please consult \cite{chataur2005adem-cartan}.

For applications related to the cohomology of spaces, it is convenient to introduce the notation $P^s = P_{-s}$ and $\beta P^s = \beta P_{-s}$, which we use in subsections 6.4 and 6.5.

\section{Effective constructions}
\label{sec: effective constructions}

In this section we construct explicit Steenrod-Adem structures on three well know combinatorial $E_\infty$-operads: the Barratt-Eccles operad $\mathcal E$ \cite{berger04combinatorial}, the surjection operad $\mathcal X$ \cite{mcclure03cochain}, and the operad $U(\mathcal M)$ associated to the finitely presented $E_\infty$-PROP $\mathcal M$ \cite{medina2020prop1}. We also define natural and effective Steenrod-Adem structure on the normalized cochains of any simplicial or cubical set, by describing a $U(\mathcal M)$-algebra structures on it.

Figure~\ref{fig: bigsummary} presents a diagrammatical representation of the constructions in this section.

\begin{figure}
	\begin{tikzcd}
	& & & U(\mathcal M) \arrow[dddr, bend left=10, "\phi^\triangle"'] \arrow[dddrr, bend left=20, "\phi^\square"'] & & \\
	& & \mathcal X \arrow[ru, "SL"] & & & \\
	& \mathcal E \arrow[ur, "TR"] & & & & \\
	\mathcal W \arrow[rrrr, "\psi^\triangle", bend right=10] \arrow[rrrrr, "\psi^\square", bend right=20] \arrow[ur, "\psi_{\mathcal E}"] \arrow[uurr, "\psi_{\mathcal X}"', bend right=35, pos=.6] \arrow[uuurrr, "\psi_{U(\mathcal M)}"', bend right=45, pos=0.6] & & & & \End_{N^\bullet(X)} & \End_{N^\bullet(X)}
	\end{tikzcd}
	\caption{Summary of effective constructions: Steenrod-Adem structures on the Barratt-Eccles $\mathcal E$, surjection $\mathcal X$, and $U(\mathcal M)$ operads, and natural Steenrod-Adem structures on the normalized chains of simplicial and cubical sets. We are abusing notation since the maps $TR$ and $SL$ required different sign conventions.}
	\label{fig: bigsummary}
\end{figure}

\subsection{Barratt-Eccles operad} In this subsection we effectively describe a Steenrod-Adem structure on the Barratt-Eccles operad.

>>>>

\section{Group homology} \label{s:goup homology}

@@
